---
title: "R Notebook"
output: html_notebook
---

# Analyze demo data for logistic regression

libraries
```{r}
library(cmdstanr)
library(posterior)
library(bayesplot)

library(ggplot2)
library(dplyr)
library(tidyverse)
```

define plotting function for stacked bar plot of observation ratio
```{r}
plot_ratiostackbar <- function(df, title="ratio-stack bar plot") {
  
  df <- na.omit(df)
  input_category <- unique(df[,1])
  output_number_positive <- integer(length(input_category))
  output_number_negative <- integer(length(input_category))
  
  for (i in 1:length(input_category)){
    df_single_category <- df[df[,1] == input_category[i],]
    output_number_positive[i] <- dim(df_single_category[df_single_category[,2] == 1,])[1]
    output_number_negative[i] <- dim(df_single_category[df_single_category[,2] == 0,])[1]
  }
  
  data_plotting <- data.frame(input_category = rep(input_category,2),
                              output_number = c(output_number_positive, output_number_negative),
                              output_class = rep(c("positive","negative"), each=length(input_category)))
  
  g <- ggplot(data_plotting, aes(fill=output_class, y=output_number, x=input_category)) + 
    geom_bar(position="fill", stat="identity") +
    ylab("observation_ratio") +
    ggtitle(label=title)
  
  plot(g)
 }
```

# demo data1 (probability range from 0% to 100%)
read demo_data1
```{r}
data <- read.delim("demoData_IR-LEGO_Cre-loxP_1.txt")
data
```
plot data overview
```{r}
data$laser_power <- as.factor(data$laser_power)
plot_ratiostackbar(df=data, title="demo data1")
```
standard logistic modeling, that is glm() with link="logit"
```{r}
data <- read.delim("demoData_IR-LEGO_Cre-loxP_1.txt")
model <- glm(recombination ~ laser_power, data=data,family=binomial(link="logit"))
summary(model)
```

```{r}
x <- seq(5, 18, length.out = 50)
y <- 1 / (1 + exp(- model$coefficients["(Intercept)"] - model$coefficients["laser_power"] * x))    # fitting model
plot_df <- data.frame(x, y)

p <- ggplot() +
  geom_line(aes(x, y), data = plot_df, size = 2, color="blue") +
  scale_x_continuous("Laser Power (mW)", limits = c(5, 20), breaks = seq(5, 20, 2),
                       ) +
    scale_y_continuous("Probability", limits = c(-0.05, 1.05),
                       breaks = seq(0, 1, 0.2))
print(p)
```


# demo data2 (not achieve 100% probability)
read demo_data2
```{r}
data <- read.delim("demoData_IR-LEGO_Cre-loxP_2.txt")
data
```
plot data overview
```{r}
data$laser_power <- as.factor(data$laser_power)
plot_ratiostackbar(df=data, title="demo data2")
```
select laser power range for a increasing function of logistic model
```{r}
power_upper_limit <- 15.5    #set upper limit of fitting data

data <- read.delim("demoData_IR-LEGO_Cre-loxP_2.txt")
data <- data[data$laser_power < power_upper_limit,] # power filtering suitable for logistic regression that is sigmoid curve
```

format the data to be input into the sampler
```{r}
data.list <- list(N=dim(data)[1], 
                        induction=data$recombination, 
                        power=data$laser_power)
print(data.list)
```
compile stan model
```{r}
file <- "IR-LEGO_creloxp_logit_plimit_model.stan" #define the model file
mod <- cmdstan_model(file)    #compile the model
```
check stan model
```{r}
mod$print()    #display the model code
```
MCMC Sampling
```{r}
### MCMC sampling for logistic model: induction ~ power
fit_plimit_model <- mod$sample(
  data = data.list,
  seed = 123,
  adapt_delta = 0.999,
  chains = 4,
  parallel_chains = 2,
  refresh = 500
)
```

Check Posterior summary statistics
```{r}
fit_plimit_model$summary()
```

```{r}
mcmc_hist(fit_plimit_model$draws("plimit"))
```

```{r}
mcmc_hist(fit_plimit_model$draws("alpha"))
```

```{r}
mcmc_hist(fit_plimit_model$draws("beta"))
```
check CmdStan’s diagnose and stansummary utilities
```{r}
fit_plimit_model$cmdstan_diagnose()
```
```{r}
fit_plimit_model$cmdstan_summary()
```
plot trace
```{r}
np_cp <- nuts_params(fit_plimit_model)
head(np_cp)
```

```{r}
#trace plot plimit
mcmc_trace(fit_plimit_model$draws("plimit")) +
  xlab("iteration, sampling phase after warming")
```
```{r}
#trace plot alpha
mcmc_trace(fit_plimit_model$draws("alpha")) +
  xlab("iteration, sampling phase after warming")
```
```{r}
#trace plot alpha
mcmc_trace(fit_plimit_model$draws("beta")) +
  xlab("iteration, sampling phase after warming")
```
format sampling results
```{r}
draws <- fit_plimit_model$draws()
fit_df <- as_draws_df(draws) #format into a data.frame
head(fit_df)     #display the sampling results
```
plot the best model from stan
```{r}
alpha <- median(fit_df$alpha)
beta <- median(fit_df$beta)
p.limit <- median(fit_df$plimit)

x <- seq(5, 15, length.out = 50)
y <- p.limit / (1 + exp(- beta - alpha * x))
plot_df <- data.frame(x, y)

p <- ggplot() +
  geom_line(aes(x, y), data = plot_df, size = 2, color="green") +
  scale_x_continuous("Laser Power (mW)", limits = c(5, 20), breaks = seq(5, 20, 2),
                       ) +
    scale_y_continuous("Probability", limits = c(-0.05, 1.05),
                       breaks = seq(0, 1, 0.2))
print(p)
```

