---
title: "logistic regression in R"
output: html_notebook
author: "Toshiaki Tameshige"
date: "2023-05-02"
---

# Analyze demo data for Bayesian logistic regression and likelihood-ratio test
#### The p-values of the likelihood-ratio test (LRT) between Bayesian models calculated in "demoData_analysis_3.Rmd" may be less accurate than those from the LRT between models based on glm function. This is because there is a concern that being non-linear model and the fitting by MCMC method may decrease the extent to which the distribution followed by the statistic, i.e. two times the log likelihood ratio or difference of deviances, approximate the chi-square distribution. The parametric bootstrap approach is a good alternative for LRT when there is a concern on the approximation by chi-square distribution. However, parametric bootstrap LRT (pbLRT) can requires a very long computational time because of the large number (>1,000) of bootstrap iterations. Here, instead of p-value calculation by pbLRT, I show how to run small number of bootstrap iterations and check the similarity between the distribution of test statistic (two times the log-likelihood ratio) and chi-square distribution under the null hypothesis.

### before run below, make sure that you have enough storage (maybe 2.2 GB) because the program below save each mcmc sampling results as RData and text files

libraries
```{r}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(tidyverse)
```

check if there is a directory "original_model_parameters" which contain the Bayesian model parameters from the original data
```{r}
list.files("original_model_parameters/")
```

some preparation for parametric bootstrap likelihood-ratio test
```{r}
# clear all objects
rm(list = ls())

# import the original fitting parameters of model2 (size-ignoring)
load("./original_model_parameters/model2_parameters.RData")

power_upper_limit <- 15    #set upper limit of fitting data
data_ori <- read.delim("demoData_IR-LEGO_Cre-loxP_3.txt")    #import data
data_ori <- data_ori[data_ori$laser_power < power_upper_limit,] # power filtering suitable for logistic regression that is sigmoid curve

data_ori$cell_size.scale <- as.numeric(scale(data_ori$cell_size))   #scaling of the cell size data
x <- data_ori$laser_power
yp <- mdl2_p.limit / (1 + exp(- mdl2_beta - mdl2_alpha * x))
```

compile the stan models
```{r}
file <- "IR-LEGO_creloxp_logit_plimitcellsize_model2.stan" #define the model file
mod1 <- cmdstan_model(file)    #compile the full model as mod1

file <- "IR-LEGO_creloxp_logit_plimit_model.stan" #define the model file
mod2 <- cmdstan_model(file)    #compile the reduced model as mod2
```

make required directories
```{r}
dir.create("mcmc_results_LR")
dir.create("mcmc_diagnose_model1")
dir.create("mcmc_diagnose_model2")
dir.create("mcmc_results_model1")
dir.create("mcmc_results_model1param")
dir.create("mcmc_results_model2")
dir.create("mcmc_results_model2param")
dir.create("mcmc_results_recombData")

```

set the trial number for bootstrap
```{r}
k_max <- 150
all_diff_dev <- numeric(k_max)
```

try loop for only 2 times (just trial)
```{r, message=FALSE}
t1 <- Sys.time()

#k <- 1
k_max <- 2

for (k in 1:k_max) {
  y01 <- integer(length(yp))
  for (i in 1:length(yp)){
    set.seed(i+k*10)
    y01[i] <- rbinom(1,1,yp[i])
  }
  
  #save the simulation data of k
  save(y01,file=paste0("mcmc_results_recombData/recombData_",
                       formatC(k, width = 4, flag = "0"),
                       ".RData"))

  ### reduced model
  # formatting the data to input into the sampler
  data.list <- list(N=dim(data_ori)[1], 
                    induction=y01, 
                    power=data_ori$laser_power)

  ### MCMC sampling for logistic model: induction ~ power
  fit_plimit_model <- mod2$sample(
    data = data.list,
    seed = 111,
    adapt_delta = 0.999,
    chains = 4,
    parallel_chains = 2,
    refresh = 500
    )

  #save the MCMC result in RData with a file name with 4-digit k number
  fit_plimit_model$save_object(file = paste0("mcmc_results_model2/fit_plimit_model_",
                                     formatC(k, width = 4, flag = "0"),
                                     ".RData"))
  
  #save the diagnose message
  msgtxt <- fit_plimit_model$cmdstan_diagnose()
  fn_dgn <- paste0("mcmc_diagnose_model2/diagnose_model2_",
                   formatC(k, width = 4, flag = "0"),
                   ".txt")
  cat(file = fn_dgn)
  writeLines(msgtxt$stdout, con = fn_dgn)

  #get the median values as the maximum likely parameters
  mdl2_alpha <- median(fit_plimit_model$draws("alpha"))
  mdl2_beta <- median(fit_plimit_model$draws("beta"))
  mdl2_p.limit <- median(fit_plimit_model$draws("plimit"))

  #save the model_2 parameters of trial k
  save(mdl2_alpha,
       mdl2_beta,
       mdl2_p.limit,
       file=paste0("mcmc_results_model2param/parameter_model2_",
                   formatC(k, width = 4, flag = "0"),
                   ".RData"))
  
  #calculate the deviance of the reduced model
  new_x <- data_ori$laser_power

  pred_prob <- mdl2_p.limit / (1 + exp(- mdl2_beta - mdl2_alpha * new_x ))   # induction rate, or probability, predicted from our model with only laser power

  pos <- (y01 == 1)
  neg <- (y01 == 0)
  #probability of unsuccess
  pred_prob_false <- 1- pred_prob

  manual_loglikelihood <- sum(log(pred_prob[pos])) + sum(log(pred_prob_false[neg]))
  manual_deviance_model2 <- -2*manual_loglikelihood    #deviance of the reduced model
  rm(manual_loglikelihood)

  ### full model
  # formatting the data to input into the sampler of full model
  data.list <- list(N=dim(data_ori)[1], 
                    induction=y01,
                    power=data_ori$laser_power, 
                    size=data_ori$cell_size.scale)

  ### MCMC sampling for logistic model: induction ~ power+size+power:size
  fit_plimitsize_model <- mod1$sample(
    data = data.list,
    seed = 111,
    adapt_delta = 0.999,
    chains = 4,
    parallel_chains = 2,
    refresh = 500
    )

  #save the MCMC result in RData with a file name with 4-digit k number
  fit_plimitsize_model$save_object(file = paste0("mcmc_results_model1/fit_plimitsize_model_",
                                         formatC(k, width = 4, flag = "0"),
                                         ".RData"))

  #save the diagnose message
  msgtxt <- fit_plimitsize_model$cmdstan_diagnose()
  fn_dgn <- paste0("mcmc_diagnose_model1/diagnose_model1_",
                   formatC(k, width = 4, flag = "0"),
                   ".txt")
  cat(file = fn_dgn)
  writeLines(msgtxt$stdout, con = fn_dgn)

  #get the median values as the maximum likely parameters
  mdl1_alpha_one <- median(fit_plimitsize_model$draws("alpha_one"))
  mdl1_alpha_two <- median(fit_plimitsize_model$draws("alpha_two"))
  mdl1_alpha_three <- median(fit_plimitsize_model$draws("alpha_three"))
  mdl1_beta <- median(fit_plimitsize_model$draws("beta"))
  mdl1_p.limit_b <- median(fit_plimitsize_model$draws("plimit_b"))
  mdl1_p.limit_a <- median(fit_plimitsize_model$draws("plimit_a"))

  #save the model_1 parameters of trial k
  save(mdl1_alpha_one,
       mdl1_alpha_two,
       mdl1_alpha_three,
       mdl1_beta,
       mdl1_p.limit_b,
       mdl1_p.limit_a,
       file=paste0("mcmc_results_model1param/parameter_model1_",
                   formatC(k, width = 4, flag = "0"),
                   ".RData"))
  
  #calculate the deviance of the full model
  new_x1 <- data_ori$laser_power
  new_x2 <- data_ori$cell_size.scale

  pred_prob <- (mdl1_p.limit_b + new_x2*mdl1_p.limit_a) / (1 + exp(- mdl1_beta - mdl1_alpha_one * new_x1 - mdl1_alpha_two * new_x2 - mdl1_alpha_three*new_x1*new_x2))   #get the induction rate, or probability, predicted from our bayesian model with laser power and cell size

  #calculate deviance by simply multiplying of probabilities

  pred_prob_false <- 1- pred_prob   #probability of unsuccess

  manual_loglikelihood <- sum(log(pred_prob[pos])) + sum(log(pred_prob_false[neg]))

  manual_deviance_model1 <- -2*manual_loglikelihood

  diff_dev <- manual_deviance_model2 - manual_deviance_model1
  
  diff_dev
  all_diff_dev[k] <- diff_dev
}
```

try fill loop for parametric bootstrap LRT
```{r, message=FALSE}
t1 <- Sys.time()

#k <- 1
k_max <- 150

for (k in 3:k_max) {
  y01 <- integer(length(yp))
  for (i in 1:length(yp)){
    set.seed(i+k*10)
    y01[i] <- rbinom(1,1,yp[i])
  }
  
  #save the simulation data of k
  save(y01,file=paste0("mcmc_results_recombData/recombData_",
                       formatC(k, width = 4, flag = "0"),
                       ".RData"))

  ### reduced model
  # formatting the data to input into the sampler
  data.list <- list(N=dim(data_ori)[1], 
                    induction=y01, 
                    power=data_ori$laser_power)

  ### MCMC sampling for logistic model: induction ~ power
  fit_plimit_model <- mod2$sample(
    data = data.list,
    seed = 111,
    adapt_delta = 0.999,
    chains = 4,
    parallel_chains = 2,
    refresh = 500
    )

  #save the MCMC result in RData with a file name with 4-digit k number
  fit_plimit_model$save_object(file = paste0("mcmc_results_model2/fit_plimit_model_",
                                     formatC(k, width = 4, flag = "0"),
                                     ".RData"))
  
  #save the diagnose message
  msgtxt <- fit_plimit_model$cmdstan_diagnose()
  fn_dgn <- paste0("mcmc_diagnose_model2/diagnose_model2_",
                   formatC(k, width = 4, flag = "0"),
                   ".txt")
  cat(file = fn_dgn)
  writeLines(msgtxt$stdout, con = fn_dgn)

  #get the median values as the maximum likely parameters
  mdl2_alpha <- median(fit_plimit_model$draws("alpha"))
  mdl2_beta <- median(fit_plimit_model$draws("beta"))
  mdl2_p.limit <- median(fit_plimit_model$draws("plimit"))

  #save the model_2 parameters of trial k
  save(mdl2_alpha,
       mdl2_beta,
       mdl2_p.limit,
       file=paste0("mcmc_results_model2param/parameter_model2_",
                   formatC(k, width = 4, flag = "0"),
                   ".RData"))
  
  #calculate the deviance of the reduced model
  new_x <- data_ori$laser_power

  pred_prob <- mdl2_p.limit / (1 + exp(- mdl2_beta - mdl2_alpha * new_x ))   # induction rate, or probability, predicted from our model with only laser power

  pos <- (y01 == 1)
  neg <- (y01 == 0)
  #probability of unsuccess
  pred_prob_false <- 1- pred_prob

  manual_loglikelihood <- sum(log(pred_prob[pos])) + sum(log(pred_prob_false[neg]))
  manual_deviance_model2 <- -2*manual_loglikelihood    #deviance of the reduced model
  rm(manual_loglikelihood)

  ### full model
  # formatting the data to input into the sampler of full model
  data.list <- list(N=dim(data_ori)[1], 
                    induction=y01,
                    power=data_ori$laser_power, 
                    size=data_ori$cell_size.scale)

  ### MCMC sampling for logistic model: induction ~ power+size+power:size
  fit_plimitsize_model <- mod1$sample(
    data = data.list,
    seed = 111,
    adapt_delta = 0.999,
    chains = 4,
    parallel_chains = 2,
    refresh = 500
    )

  #save the MCMC result in RData with a file name with 4-digit k number
  fit_plimitsize_model$save_object(file = paste0("mcmc_results_model1/fit_plimitsize_model_",
                                         formatC(k, width = 4, flag = "0"),
                                         ".RData"))

  #save the diagnose message
  msgtxt <- fit_plimitsize_model$cmdstan_diagnose()
  fn_dgn <- paste0("mcmc_diagnose_model1/diagnose_model1_",
                   formatC(k, width = 4, flag = "0"),
                   ".txt")
  cat(file = fn_dgn)
  writeLines(msgtxt$stdout, con = fn_dgn)

  #get the median values as the maximum likely parameters
  mdl1_alpha_one <- median(fit_plimitsize_model$draws("alpha_one"))
  mdl1_alpha_two <- median(fit_plimitsize_model$draws("alpha_two"))
  mdl1_alpha_three <- median(fit_plimitsize_model$draws("alpha_three"))
  mdl1_beta <- median(fit_plimitsize_model$draws("beta"))
  mdl1_p.limit_b <- median(fit_plimitsize_model$draws("plimit_b"))
  mdl1_p.limit_a <- median(fit_plimitsize_model$draws("plimit_a"))

  #save the model_1 parameters of trial k
  save(mdl1_alpha_one,
       mdl1_alpha_two,
       mdl1_alpha_three,
       mdl1_beta,
       mdl1_p.limit_b,
       mdl1_p.limit_a,
       file=paste0("mcmc_results_model1param/parameter_model1_",
                   formatC(k, width = 4, flag = "0"),
                   ".RData"))
  
  #calculate the deviance of the full model
  new_x1 <- data_ori$laser_power
  new_x2 <- data_ori$cell_size.scale

  pred_prob <- (mdl1_p.limit_b + new_x2*mdl1_p.limit_a) / (1 + exp(- mdl1_beta - mdl1_alpha_one * new_x1 - mdl1_alpha_two * new_x2 - mdl1_alpha_three*new_x1*new_x2))   #get the induction rate, or probability, predicted from our bayesian model with laser power and cell size

  #calculate deviance by simply multiplying of probabilities

  pred_prob_false <- 1- pred_prob   #probability of unsuccess

  manual_loglikelihood <- sum(log(pred_prob[pos])) + sum(log(pred_prob_false[neg]))

  manual_deviance_model1 <- -2*manual_loglikelihood

  diff_dev <- manual_deviance_model2 - manual_deviance_model1
  
  diff_dev
  all_diff_dev[k] <- diff_dev
}

t2 <- Sys.time()
```

see the time it took in the program above
```{r}
print(t2 - t1)
```

check the obtained test statistics (difference of deviances under null hypothesis)
```{r}
all_diff_dev
```

save check the obtained test statistics
```{r}
save(all_diff_dev, file=paste0("mcmc_results_LR/all_diff_dev_",
                                         formatC(k, width = 4, flag = "0"),
                                         ".RData"))
```

Check the distribution. 
Note that sometimes it includes minus values which are results from incorrect MCMC sampling. They should be removed.
```{r}
all_diff_dev.plus <- all_diff_dev[all_diff_dev > 0]
hist(all_diff_dev.plus,main="histogram of bootstrap")
```
### compare the bootstrap and theoretical distribution
```{r}
histogram <- hist(all_diff_dev.plus, 
                  breaks=seq(from=0, to=13.5, by=1), 
                  ylim=c(0,.25), 
                  freq=FALSE,
                  main="histogram of bootstrap and theoretical chi-square distribution")

LR <- seq(from=0, to=13.5, by=0.5)
d_theo <- dchisq(LR,df=3)
lines(x=LR, 
      y=d_theo, # density
      lwd = 2, # thickness of line
      col = "gray")
```

### compare cumulative distribution
get cumulative distribution at some points of the distribution of the bootstrap
```{r}
LR <- seq(from=0, to=7.5, by=0.5)

q_bootstrap <- numeric(length(LR))
for(i in 1:length(LR)){
  q_bootstrap[i] <- 1 - (length(all_diff_dev.plus[all_diff_dev.plus> LR[i]])/length(all_diff_dev.plus))
}

#bootstrap distribution
q_bootstrap
```

get cumulative distribution at some points of the theoretical distribution
```{r}
#theoretical distribution
q_theo <- pchisq(LR,df=3)
q_theo
```

compare the two cumulative distributions by QQ-plotting
```{r}
plot(q_bootstrap,q_theo, 
     main="QQ-plot of bootstrap and theoretical chi-square")
abline(0,1,col="red",lty=2)
```
As shown in the plots above, the distributions of bootstrap and theoretical chi-square are similar enough so that p-value calculation based on chi-square distribution is a good approximation.
