---
title: "logistic regression in R"
output: html_notebook
author: "Toshiaki Tameshige"
date: "2022-11-22"
---

# Analyze demo data for logistic regression and likelihood-ratio test

libraries
```{r}
library(cmdstanr)
library(posterior)
library(bayesplot)
library(tidyverse)
```

define plotting function for stacked bar plot of observation ratio
```{r}
plot_ratiostackbar <- function(df, title="ratio-stack bar plot") {
  
  df <- na.omit(df)
  input_category <- unique(df[,1])
  output_number_positive <- integer(length(input_category))
  output_number_negative <- integer(length(input_category))
  
  for (i in 1:length(input_category)){
    df_single_category <- df[df[,1] == input_category[i],]
    output_number_positive[i] <- dim(df_single_category[df_single_category[,2] == 1,])[1]
    output_number_negative[i] <- dim(df_single_category[df_single_category[,2] == 0,])[1]
  }
  
  data_plotting <- data.frame(input_category = rep(input_category,2),
                              output_number = c(output_number_positive, output_number_negative),
                              output_class = rep(c("positive","negative"), each=length(input_category)))
  
  g <- ggplot(data_plotting, aes(fill=output_class, y=output_number, x=input_category)) + 
    geom_bar(position="fill", stat="identity") +
    ylab("observation_ratio") +
    ggtitle(label=title)
  
  return(g)
 }
```

# demo data4 (probability range from 0% to 100%)
read demo_data4
```{r}
data <- read.delim("demoData_IR-LEGO_Cre-loxP_4.txt")
data
```
plot data overview
```{r}
data$laser_power <- as.factor(data$laser_power)
df <- data[,c("laser_power","recombination")]
g <- plot_ratiostackbar(df=df, title="demo data4")+xlab("laser power (mW)")
print(g)
```
plot the cell size distribution
```{r}
data %>% ggplot(aes(x = laser_power, y = cell_size)) +
    geom_jitter(position = position_jitter(height = 0, width = 0.1), size=0.5) +
  xlab("laser power (mW)") +
  ylab(expression(paste ("cell size (", {µm^2}, ")", sep="")))+
  ggtitle("demo data4")
```

standard logistic modeling, that is glm() with link="logit"
here we assume cell-size effects on "intercept" (independent effect) and "slope" (interaction with laser_power)
```{r}
data <- read.delim("demoData_IR-LEGO_Cre-loxP_4.txt")    #import data
model <- glm(recombination ~ laser_power + cell_size + laser_power:cell_size, data=data,family=binomial(link="logit"))
summary(model)
```
plot the curves of the model at some cell sizes
```{r}
cell_size_4plot <- c(2400,2800,3200,3600,4000)
n_xpoints <- 50

x <- seq(5, 16, length.out = n_xpoints)

plot_df <- data.frame(x=numeric(0),y=numeric(0),cell_size=numeric(0))

for (i in 1:length(cell_size_4plot)){
  y <- 1 / (1 + exp(- model$coefficients["(Intercept)"] - model$coefficients["laser_power"] * x - model$coefficients["cell_size"] * cell_size_4plot[i] - model$coefficients["laser_power:cell_size"] *x* cell_size_4plot[i]))    # simulate y according to the model
  each_df <- data.frame(x, y,cell_size=rep(cell_size_4plot[i],times=n_xpoints))
  plot_df <- rbind(plot_df,each_df)
}
plot_df$cell_size <- as.factor(plot_df$cell_size)

#plot with variation in cell size
p <- ggplot(data = plot_df) +
  geom_line(aes(x=x, y=y, color=cell_size), size = 1.5) +
  scale_x_continuous("Power (mW)", limits = c(4, 16), breaks = seq(4, 16, 2),
                       ) +
    scale_y_continuous("Pr (recombination)", limits = c(-0.05, 1.05),
                       breaks = seq(0, 1, 0.2))
print(p)
```
likelihood-ratio test
```{r}
library(lmtest)

#model1 is a reduced model
model1 <- glm(recombination ~ laser_power, data=data,family=binomial(link="logit"))
# model2 is a full model
model2 <- glm(recombination ~ laser_power + cell_size + laser_power:cell_size, data=data,family=binomial(link="logit"))

#likelihood-ratio test
lrtest(model2,model1)
```
The cell_size effect is significant!



# demo data3 (not achieving 100% probability)
read demo_data3
```{r}
data <- read.delim("demoData_IR-LEGO_Cre-loxP_3.txt")
data
```
plot data overview
```{r}
data$laser_power <- as.factor(data$laser_power)
df <- data[,c("laser_power","recombination")]
g <- plot_ratiostackbar(df=df, title="demo data3")+xlab("laser power (mW)")
print(g)
```
select laser power range for a increasing function of logistic model
```{r}
power_upper_limit <- 15    #set upper limit of fitting data

data <- read.delim("demoData_IR-LEGO_Cre-loxP_3.txt")    #import data
data <- data[data$laser_power < power_upper_limit,] # power filtering suitable for logistic regression that is sigmoid curve

#scaling of the cell size data
data$cell_size.scale <- as.numeric(scale(data$cell_size))

dim(data)
```

format the data to be input into the sampler
```{r}
data.list <- list(N=dim(data)[1], 
                        induction=data$recombination, 
                        power=data$laser_power, 
                        size=data$cell_size.scale)
print(data.list)
```
compile stan model
```{r}
file <- "IR-LEGO_creloxp_logit_plimitcellsize_model.stan" #define the model file
mod <- cmdstan_model(file)    #compile the model
```
check stan model
```{r}
mod$print()    #display the model code
```
MCMC Sampling
```{r}
### MCMC sampling for logistic model: induction ~ power+size+power:size
fit_plimitsize_model <- mod$sample(
  data = data.list,
  seed = 111,
  adapt_delta = 0.999,
  chains = 4,
  parallel_chains = 2,
  refresh = 500
)
```
check diagnostics
```{r}
fit_plimitsize_model$cmdstan_diagnose()
```
check sampling summary
```{r}
fit_plimitsize_model$cmdstan_summary()
```
trace plot
```{r}
#trace plot plimit_a
mcmc_trace(fit_plimitsize_model$draws("plimit_a")) +
  xlab("iteration, only post-warmup")
```

sampled posterior distribution
```{r}
mcmc_hist(fit_plimitsize_model$draws("plimit_a"))
```

```{r}
mcmc_hist(fit_plimitsize_model$draws("plimit_b"))
mcmc_hist(fit_plimitsize_model$draws("alpha_one"))
mcmc_hist(fit_plimitsize_model$draws("alpha_two"))
mcmc_hist(fit_plimitsize_model$draws("alpha_three"))
mcmc_hist(fit_plimitsize_model$draws("beta"))
```

Now, we adopt the median values of parameter samples because some distribution shows skewness making the mean values not the best representative.
get median values from the samples as the maximum likelihood parameters
```{r}
alpha_one <- median(fit_plimitsize_model$draws("alpha_one"))
alpha_two <- median(fit_plimitsize_model$draws("alpha_two"))
alpha_three <- median(fit_plimitsize_model$draws("alpha_three"))
beta <- median(fit_plimitsize_model$draws("beta"))
p.limit_b <- median(fit_plimitsize_model$draws("plimit_b"))
p.limit_a <- median(fit_plimitsize_model$draws("plimit_a"))
```

plot the model at some cell sizes: 2400,2800,3200,3600,4000
```{r}
cell_size_4plot <- c(2400,2800,3200,3600,4000)

mean_cell_size_raw <- mean(data$cell_size)
sd_cell_size_raw <- sd(data$cell_size)
cell_size_4plot.scale <- (cell_size_4plot - mean_cell_size_raw)/sd_cell_size_raw    #scale the cell size

n_xpoints <- 30
plot_df <- data.frame(x=numeric(0),y=numeric(0),cell_size=numeric(0))

for (i in 1:length(cell_size_4plot)){
  x <- seq(5.5, 14, length.out = n_xpoints)
  y <- (p.limit_b + cell_size_4plot.scale[i]*p.limit_a) / (1 + exp(- beta - alpha_one * x - alpha_two * cell_size_4plot.scale[i] - alpha_three*x*cell_size_4plot.scale[i]))
  each_df <- data.frame(x, y,cell_size=rep(cell_size_4plot[i],times=n_xpoints))
  plot_df <- rbind(plot_df,each_df)
}
plot_df$cell_size <- as.factor(plot_df$cell_size)

p <- ggplot(data = plot_df) +
  geom_line(aes(x=x, y=y, color=cell_size), size = 1.5) +
  scale_x_continuous("Power (mW)", limits = c(4, 20), breaks = seq(4, 20, 2),
                       ) +
    scale_y_continuous("Pr (recombination)", limits = c(-0.05, 1.05),
                       breaks = seq(0, 1, 0.2))
print(p)
```

next, calcurate deviance of the models
full model with cell_size, cell_size-laser_power interaction and cell_size-p.limit interaction

```{r}
new_x1 <- data$laser_power
new_x2 <- data$cell_size.scale

pred_prob <- (p.limit_b + new_x2*p.limit_a) / (1 + exp(- beta - alpha_one * new_x1 - alpha_two * new_x2 - alpha_three*new_x1*new_x2))   #get the induction rate, or probability, predicted from our bayesian model with laser power and cell size

#calculate deviance by simply multiplying of probabilities
pos <- (data$recombination == 1)
neg <- (data$recombination == 0)

#probability of unsuccess
pred_prob_false <- 1- pred_prob

manual_loglikelihood <- sum(log(pred_prob[pos])) + sum(log(pred_prob_false[neg]))

manual_deviance_model1 <- -2*manual_loglikelihood
paste0("full model deviance = ",manual_deviance_model1)
```

next, a reduced model ignoring cell-size effect
format the data to be input into the sampler

```{r}
data.list <- list(N=dim(data)[1], 
                        induction=data$recombination, 
                        power=data$laser_power)
print(data.list)
```
compile stan model
```{r}
file <- "IR-LEGO_creloxp_logit_plimit_model.stan" #define the model file
mod <- cmdstan_model(file)    #compile the model
```
check stan model
```{r}
mod$print()    #display the model code
```
```{r}
### MCMC sampling for logistic model: induction ~ power
fit_plimit_model <- mod$sample(
  data = data.list,
  seed = 111,
  adapt_delta = 0.999,
  chains = 4,
  parallel_chains = 2,
  refresh = 500
)
```
check the posterior
```{r}
mcmc_hist(fit_plimit_model$draws("plimit"))
mcmc_hist(fit_plimit_model$draws("alpha"))
mcmc_hist(fit_plimit_model$draws("beta"))
```
get median values from the samples as the maximum likelihood parameters
```{r}
alpha <- median(fit_plimit_model$draws("alpha"))
beta <- median(fit_plimit_model$draws("beta"))
p.limit <- median(fit_plimit_model$draws("plimit"))
```
plot the resulting model
```{r}
x <- seq(5.5, 14, length.out = 30)
y <- p.limit / (1 + exp(- beta - alpha * x))
plot_df <- data.frame(x, y)

p <- ggplot() +
  geom_line(aes(x, y), data = plot_df, size = 2, color="red") +
  
  scale_x_continuous("Power (mW)", limits = c(4, 20), breaks = seq(4, 20, 2),
                       ) +
    scale_y_continuous("Pr (recombination)", limits = c(-0.05, 1.05),
                       breaks = seq(0, 1, 0.2))
print(p)
```

```{r}
new_x <- data$laser_power

pred_prob <- p.limit / (1 + exp(- beta - alpha * new_x ))   # induction rate, or probability, predicted from our model with only laser power

pos <- (data$recombination == 1)
neg <- (data$recombination == 0)

#probability of unsuccess
pred_prob_false <- 1- pred_prob

manual_loglikelihood <- sum(log(pred_prob[pos])) + sum(log(pred_prob_false[neg]))

manual_deviance_model2 <- -2*manual_loglikelihood
paste0("cell size-ignored model deviance = ",manual_deviance_model2)
```

Then difference of deviance is
```{r}
diff_dev <- manual_deviance_model2 - manual_deviance_model1
print(diff_dev)
```
What is chisq value at df=3, p<0.05?
```{r}
qchisq(0.05,3,lower.tail = FALSE)
```
the p-value of the likelihood-raio test of our bayesian models
```{r}
p_LRT <- pchisq(diff_dev, 3,lower.tail = FALSE)
paste0("likelihood-ratio test, p = ",p_LRT)
```
The cell_size effect is significant!

```{r}
sessionInfo()
```

